---
title: "Running LASSO regressions"
author: "Anna Elisabeth Furtjes"
highlighter: prettify
output: 
  html_document:
    code_folding: hide
    theme: spacelab
    includes:
      after_body: footer.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```


<br>
<br>


***

Here, we perform LASSO models to obtain ROI-specific effect sizes that can be used to predict individual-level non-brain traits in an independent prediction dataset. 
We obtain absolute errors as the difference between observed and predicted values for a non-brain trait. 
To test whether one atlas yielded better prediction accuracy than another, we tested whether there was a statistical difference in absolute errors between two atlases using the Wilcoxon Sign Rank test. 



<br>
<br>


# LASSO regression 

The code below is an R script (`LASSO_bigstatsr.R`) we use to calculate LASSO models and from which we get individual-level prediction into the phenotype. This script outputs text files with participant IIDs (validation only) and predicted phenotypic values. 


```{r}
### Calculate LASSO models
### Script adapted from 
### https://github.com/baptisteCD/Brain-LMM/blob/master/RR_9.0_PenalisedRegressionFunction_bigstatsr.R
### 250GB RAM
# Load environment
library(readr)
library(bigmemory)
library(doMC)
library(bigstatsr)
library(stringr)


# Load arguments
arg = commandArgs(trailingOnly=TRUE)

# Arguments are
# 1) working directory - where files are located
# 2) file of vertex wise values/ or atlas file name
# 3) output folder (in working directory)
# 4) variable (to predict) name
#### calling it in shell with eg: Rscript LASSO_bigstatsr.R /working/directory /atlases/JulichBrain_ThickAvg $pheno


wd = arg[1]

vertexFile = arg[2] ## bod file processed through OSCA


pheno = arg[3]

outFolder = arg[4]


# Working directory
setwd(wd)

print("Finished set-up.")

# Big statsr approach - attache previously created FBM
st=Sys.time()


dat=read.big.matrix(paste0(wd,"/", vertexFile), sep = " ", header = T, type="double") 

et=Sys.time()
openingTime=et-st

# Write time
print(paste("Finished reading in brain data."))
openingTime

# Check dimensions
print(dim(dat))
nfeatures = ncol(dat)

# Open phenotype and order according to the ids list
phenoVar = read.table(paste0("~/phenotypes/", pheno, ".txt"), sep = "\t", header = F)
names(phenoVar) = c("IID","FID","pheno")

print(paste("Finished reading in pheno data."))


# create table that holds info on this analysis 
log = data.frame(matrix(ncol=2,nrow=11))
names(log) = c("measure", "info")

####################################################################
#### Define discovery and replication data set
####################################################################

## implement random sampling in script
all_brain_subj = as.data.frame(dat[,1])
names(all_brain_subj)=c("IID")

all_subj = merge(phenoVar, all_brain_subj, by="IID")

## restrict to participants who were included in the previous morph estimates 
print("Reading in restrict table")
restrict = read.table("~/morphometricity/data/indi.list.fid")
names(restrict)=c("IID","FID")
all_subj = merge(all_subj, restrict, by="IID")

# only keep unique participants (no duplicates)
all_subj = unique(all_subj$IID)


# randomly sample
set.seed(12345)
disco = sample(all_subj, size = 0.8*length(all_subj), replace=FALSE)
# none of these participants have missing phenotypes as there are no missing values in my phenos
repli = setdiff(all_subj, disco)

print(paste0("This analysis is for atlas ", vertexFile, " and phenotype: ",pheno,", including ", length(disco)," participants in the discovery data set, and ", length(repli), " in the replication data set (Ntotal = ",length(all_subj),")."))

# save this info in log object
log[1,] =c("atlas: ",vertexFile)
log[2,] =c("pheno: ",pheno)
log[3,] =c("Ndisco: ",length(disco))
log[4,] =c("Nrepli: ",length(repli))
log[5,] =c("Ntotal: ",length(all_subj))


print(paste("Finished splitting sample into discovery and replication."))



####################################################################
#### LASSO 
####################################################################

print("Starting to apply split to big matrix")
# split brain data set
datD = dat[which(dat[,1] %in% disco),]
datR = dat[which(dat[,1] %in% repli),]
print("Finished applying split to big matrix")

# Check dimensions of subsets
print(dim(datD))
print(str(datD))

print(dim(datR))

print("Transforming as_FBM")
# Format for bigstatsr
datD_FBM = as_FBM(datD) 
datR_FBM = as_FBM(datR) 
print("Finished transforming as_FBM")


print(pheno)
print("Opening phenotype file")

# Re order and match IDs with phenotypes
phenoM=phenoVar[match(datD[,1], phenoVar[,1], nomatch = NULL),c("IID","pheno")]

# check that IDs from brain & pheno file match
cor(datD[,1], phenoM[,1])
cbind(datD[1:10,1], phenoM[1:10,1])

print("Cross validation using bigstatsr (NB not real CV)")
st1=Sys.time()

# Specifications:
# X = An object of class FBM. (brain pheno)
# y.train = Vector of responses (behavioral pheno)
# ind.col = Vector of the column indices that are used (all but IID column, brain and pheno frames should have same order)
# alphas = lasso penalty is 1 
# nlambda = The number of lambda values. Default is 200
# K = Number of sets used in the Cross-Model Selection and Averaging (CMSA) procedure. (ten-fold cross-valid)

print(paste("Now starting ten-fold cross-validation."))


cv.bstatsr <- big_spLinReg(X = datD_FBM, y.train = phenoM[,2], ind.col = 2:nfeatures, alphas = 1, nlambda = 100, K = 10)
et1=Sys.time()
cvRunTime=et1-st1
# Write time
print(paste("Finished running cross-validation using bigstatsr"))
cvRunTime

# Print lasso details
summary(cv.bstatsr)

# save this info in log object
log[6,] =c("nfeatures: ",summary(cv.bstatsr)$nb_var)
log[7,] =c("alpha: ",summary(cv.bstatsr)$alpha)
log[8,] =c("power_adaptive: ",summary(cv.bstatsr)$power_adaptive)
log[9,] =c("power_scale: ",summary(cv.bstatsr)$power_scale)
log[10,] =c("validation_loss: ",summary(cv.bstatsr)$validation_loss)
log[11,] =c("intercept: ",summary(cv.bstatsr)$intercept)

#log[12:(11+summary(cv.bstatsr)$nb_var),"measure"]=paste0("beta",1:summary(cv.bstatsr)$nb_var)
#log[12:(11+summary(cv.bstatsr)$nb_var),"info"]=unlist(summary(cv.bstatsr)$beta)

print(paste("Stored results in log."))


# Store predictor
# this should have nrow = repli
res=cbind(datR[,1], as.vector(predict(cv.bstatsr, datR_FBM )) )

# get atlas name to print
atlas=strsplit(vertexFile, "_")[[1]][1]
meas=strsplit(vertexFile, "_")[[1]][3]

write.table(res, paste0(outFolder, "/", "lasso", "_pred_", atlas,"_",meas, "_", pheno,".txt"), col.names = F, row.names = F, quote=F)
write.table(log, paste0(outFolder, "/", "lasso", "_pred_", atlas,"_",meas, "_", pheno,".log"), col.names = F, row.names = F, quote=F)

print(paste("Saved output in ",outFolder))

```


<br>



The code below is used to initiate the Rscript above.

```{bash}
#!/bin/bash -l
#SBATCH --job-name=LASSO_atlases
#SBATCH --output=~/morphometricity/output/LASSO_atlases_%A_%a.out
#SBATCH -D ~/imaging/output
#SBATCH --mem=100G 
#SBATCH --cpus-per-task=10
#SBATCH --time=2-00:00
#SBATCH --array=1-7 ## one per meas

wd="~/morphometricity/data/atlases"
bod_dir="~/morphometricity/data/bod_files"
out_dir="~/morphometricity/results/LASSO"
scripts_dir="~/morphometricity/scripts"


refList=${scripts_dir}/ref_atlases.txt
atlas=$(awk "FNR==$SLURM_ARRAY_TASK_ID" $refList)


module load apps/R/3.6.0

for meas in ThickAvg SurfArea GrayVol
do

for pheno in sex_pheno bmi_pheno edu_pheno alc_pheno cig_pheno gpheno age_pheno
do

	if [ ! -f ${out_dir}/lasso_pred_${atlas}_${meas}_${pheno}.txt ]
	then
	
		
		atlasFile=${wd}/${atlas}_${meas}_*
		
		echo "Now starting LASSO model for ${atlasFile} and $pheno"
		
		Rscript ${scripts_dir}/LASSO_bigstatsr.R ${atlasFile} $pheno $out_dir
	fi
	
done

done

#### sbatch -p brc,shared ~/morphometricity/scripts/Run_LASSO_atlases.sh
```


<br>
<br>


***



# Wilcoxon Sign Ranked tests

The Rscript below (`Wilcoxon_sign_rank.R`) reads in the output obtained from the LASSO models above, calculates absolute errors as the difference between observed and predicted values, and calculates a Wilcoxon sigh rank test to compare prediction accuracies of one atlas versus another atlas. 

  - *H0* = there is no median difference in absolute error between atlas1 and atlas2 (i.e., `atlas1$abs_err - atlas2$abs_err = 0`)
  
  - *H1* = there is a difference in absolute error and it's either positive or negative (two.sided test)


A non-significant result with a small effect size tells us that we have no evidence to conclude that atlas 1 performed any better than atlas 2. 


```{r}
## Wilcoxon Signed Rank test to compare prediction accuracies obtained from different atlases 

# Analysis steps
# 1. Read in output created by LASSO predictions, eg., "lasso_pred_Yeo_GrayVol_sex.txt"
# 2. Merge predictions with actual measured phenotypes. Resulting table will have columns: eid, pheno, LASSOpredictions. End up with subset of replication sample
# 3. Calculate absolute errors: observed - averaged predictions
# 4. Perform wilcoxon test to contrast absolute errors by two atlases = absolute error atlas 1 vs. absolute error atlas2

# Load arguments
arg = commandArgs(trailingOnly=TRUE)

lasso_dir = arg[1]
# eg lasso_dir = "~/morphometricity/results/LASSO"

atlas1_name = arg[2]
# eg. atlas1_name = "Yeo"

atlas2_name = arg[3]
#eg. atlas2_name = "DK"

meas = arg[4]
# eg. meas = "ThickAvg"

pheno_name = arg[5]
# eg. pheno_name = "bmi_pheno"


# load dependencies
library(data.table)
library(stringr)
library(rstatix)
library(reshape2)


########################################################################
## read in LASSO predictions
########################################################################

# read in prediction atlas 1
atlas1_file = list.files(path = lasso_dir, pattern = paste0("lasso_pred_", atlas1_name, "_", meas, "_", pheno_name, ".txt"))
atlas1_pred = fread(paste0(lasso_dir, "/", atlas1_file), header=F)

names(atlas1_pred) = c("IID", "pred")

# read in prediction atlas 2
# if vertices, change the way atlas is referenced

if(atlas2_name == "vertices"){

	# change meas naming
	if(meas == "GrayVol"){meas_vertex = "volume"}
	if(meas == "ThickAvg"){meas_vertex = "thickness"}
	if(meas == "SurfArea"){meas_vertex = "area"}
	
	atlas2_file = list.files(path = lasso_dir, pattern=paste0("lasso_pred_vertices_", meas_vertex,".txt_",pheno_name,".txt"))
	atlas2_pred = fread(paste0(lasso_dir, "/", atlas2_file), header=F)
	
}else{
	atlas2_file = list.files(path = lasso_dir, pattern = paste0("lasso_pred_", atlas2_name, "_", meas, "_", pheno_name, ".txt"))
	atlas2_pred = fread(paste0(lasso_dir, "/", atlas2_file), header=F)
}
names(atlas2_pred) = c("IID", "pred")

########################################################################
## read in actual observed phenotype
########################################################################

pheno_dir = "/mnt/lustre/datasets/ukbiobank/ukb40933/phenotypes"

# read in atlas 1
pheno_name.txt = list.files(path = pheno_dir, pattern=paste0(pheno_name, ".txt"))
pheno = fread(paste0(pheno_dir, "/", pheno_name.txt))

# keep only one IID column
pheno = pheno[,2:3]

# name columns
names(pheno) = c("IID", "obs")


########################################################################
## merge observed and predicted
########################################################################

# merge predicted and observed values, this function will only retain overlap (i.e., replication dataset)
atlas1 = merge(atlas1_pred, pheno, by = "IID")
atlas2 = merge(atlas2_pred, pheno, by = "IID")

########################################################################
## calculate absolute errors
## difference between predicted and observed measures 
########################################################################

atlas1$abs_err1 = atlas1$pred - atlas1$obs
atlas2$abs_err2 = atlas2$pred - atlas2$obs


########################################################################
## reshape data 
########################################################################

both = merge(atlas1, atlas2, by = "IID")

both = melt(both[,c("IID", "abs_err1", "abs_err2")], id.vars = c("IID"))
names(both)=c("IID", "atlas", "abs_err")
both$atlas = ifelse(both$atlas == "abs_err1", "atlas1", "atlas2")

########################################################################
## calculate Wilcoxon Sign rank test: observed and predicted
########################################################################
## H0 = there is no difference in absolute error in the difference between observed & predicted values (i.e., atlas1$abs_err - atlas2$abs_err = 0)
## H1 = There is a difference, the median difference is positive or negative (two.sided test)
## The test statistic for the Wilcoxon Signed Rank Test is W, defined as the smaller of W+ (sum of the positive ranks) and W- (sum of the negative ranks). If the null hypothesis is true, we expect to see similar numbers of lower and higher ranks that are both positive and negative (i.e., W+ and W- would be similar). If the research hypothesis is true we expect to see more higher and positive ranks (in this example, more children with substantial improvement in repetitive behavior after treatment as compared to before, i.e., W+ much larger than W-). https://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/BS704_Nonparametric/BS704_Nonparametric6.html?msclkid=62a38d40c49b11ecb22117b019c6db2b

test = wilcox_test(data = both, abs_err ~ atlas, paired = TRUE, p.adjust.method = "bonferroni", alternative = "two.sided", detailed=TRUE)
eff_size = wilcox_effsize(data = both, abs_err ~ atlas, paired = TRUE, p.adjust.method = "bonferroni", alternative = "two.sided", detailed=TRUE)
 
# create frame to store data
results = data.frame(matrix(ncol = 8, nrow = 1))
names(results) = c("atlas1", "atlas2","meas","pheno", "N", "statistic", "p", "effsize")

# store data 
results[1,] = as.vector(c(atlas1_name, atlas2_name, meas, str_remove(pheno_name, pattern=".txt"), nrow(atlas1), test$statistic, test$p, eff_size$effsize))


## a non-significant results with a small effect size tells us that we don;t have evidence to conclude that atlas 1 performed any better than atlas 2



write.table(results, paste0("~/morphometricity/results/Wilcoxon/", atlas1_name, "_", atlas2_name, "_", meas, "_", str_remove(pheno_name, pattern=".txt"), "_wilcoxon_results.txt"), quote=F, col.names=T, row.names=F, sep="\t")
```


The Rscript above (`Wilcoxon_sign_rank.R`) can be initiated by the following shell script. 

```{bash}
#!/bin/bash -l
#SBATCH --job-name=Wilcoxon
#SBATCH --output=~/morphometricity/output/Wilcoxon_atlases_%A_%a.out
#SBATCH -D ~/imaging/output
#SBATCH --mem=10G 
#SBATCH --time=2-00:00
#SBATCH --array=1-7 ## one per meas

lasso_dir="~/morphometricity/results/LASSO"
scripts_dir="~/morphometricity/scripts"
results_dir="~/morphometricity/results/Wilcoxon"

refList=${scripts_dir}/ref_atlases.txt
atlas1=$(awk "FNR==$SLURM_ARRAY_TASK_ID" $refList)

atlas2_num=$(($SLURM_ARRAY_TASK_ID+1))

module load apps/R/3.6.0

for atlas2_find in $(seq $atlas2_num 8)
do
refList2=${scripts_dir}/ref_brain_rep.txt
atlas2=$(awk "FNR==$atlas2_find" $refList2)

	for meas in ThickAvg SurfArea GrayVol
	do

	for pheno in sex_pheno bmi_pheno edu_pheno alc_pheno cig_pheno gpheno age_pheno
	do

		if [ ! -f ${results_dir}/${atlas1}_${atlas2}_${pheno}_wilcoxon_results.txt ]
		then
		
						
			echo "Now starting Wilcoxon test for ${atlas1}_${atlas2}_${pheno}"
			
			Rscript ${scripts_dir}/Wilcoxon_sign_rank.R ${lasso_dir} ${atlas1} ${atlas2} ${meas} ${pheno}
		fi
		
	done

	done

done

#### sbatch -p brc,shared ~/morphometricity/scripts/Wilcoxon.sh
```




<br>


Format output by these scripts using the following Rscript. 

```{r}
##### format Wilcoxon results 

setwd("~/morphometricity/results/Wilcoxon")

# create object to hold results
results = data.frame(matrix(ncol = 8, nrow = 0))
names(results) = c("atlas1", "atlas2","meas","pheno", "N", "statistic", "p", "effsize")


output = list.files(pattern="wilcoxon_results.txt")
### expecting (28*3*7) results
for(i in 1:length(output)){
	line = read.table(output[i], header=T)
	
	results = rbind(results, line)

}

print(paste("Obtained results for", nrow(results), "Wilcoxon models."))

write.table(results, "~/morphometricity/results/Wilcoxon/Wilcoxon_results.table", col.names=T, row.names=T, quote=F, sep="\t")
```




<br>

***

